Modelling Stochastic Time Delay for Regression

### 2021-04-05 documentation/article/article.tex
improve authors section
improve methodology notation

\title{Modelling Stochastic Time Delay for Regression}
\author{Aaron Pickering, Dr. Juan Camilo Orduz}
%\address{}
\email{aaron1rcl@gmail.com}
%\urladdr{} 
\date{\today}
\maketitle

@@

% TODO: Rewrite concrete expression of support...is it on x ot t? -- Aaron: f(x(t)). Consider x(t) = 0.1 for t is odd, 1 for t is even. If f(x(t)) = 0 for t is odd, 1 for t is even then the series is still valid for our method.  
Let $x: \mathcal{D}_n\subset [a, b] \longrightarrow \mathbb{R}$, where $\mathcal{D}_n$ is a finite set of $n$ points. We want to determine the, functional relationship $f:\mathbb{R} \longrightarrow \mathbb{R}$ to some other (target) variable $y$. Furthermore, we impose the condition that $f$ be injective.  The series $x(t)$ must be stationary with a known value for the point at which $f(x(t)) = 0$.  That is, we know at which point the input series has no effect on the output (when it’s off). The size of the support of $f(x(t))$ should be small relative to the total vector length (we we will explore this condition further in the 'Limitations' section). We utilise the terminology {\em impulse} for an individual element of the support of $f$.

@@

\begin{align}
x(t) = 
\left(
\begin{array}{cccccc}
0 & 0 & 1 & 0 & 1 & 0
\end{array}  
\right)
\end{align}

@@

\begin{align}
X(t) = 
\left(
\begin{array}{cccccc}
0 & 0 & 1 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 1 & 0 
\end{array}  
\right)
\end{align}

@@

% TODO fix notation for T
\begin{align}
y(t)= f(1^TX(t + \mathcal{T})) + \varepsilon
\end{align}

where $X(t + \mathcal{T})$ is the matrix of time shifted impulses, 
 $X(t)$ is the observed input, 
 $1$ is the vector $(1,1,1...1)$,
 $\varepsilon$ is a noise term and $\mathcal{T}$ is the set of single impulse time delays ($\tau$'s).\\

We assume that $\mathcal{T}$ is not a constant, but rather a random draw from some discrete distribution (e.g. discrete gaussian, poisson etc). For the discrete gaussian kernel $\mathcal{T}\sim T(\mu, \sigma)$ or for the poisson distribution $\mathcal{T}\sim Pois(\lambda)$ and $\varepsilon \sim N(\mu, \sigma_{\varepsilon})$.

In order to find the best estimate of $f$, we would like to find the function $f$ which maximises the joint likelihood of the time-domain shifts and the prediction residuals. Specifically, we maximise:
% TODO: Clean this notation.
\begin{align}
    \sum_{i=1}^{n} \log(p(y_i | X_{\tau i} = x_i; \theta_{f}, \tau_i)) + \sum_{i=1}^{n} \log(p(\tau_i| \theta_{\mathcal{T}}))
\end{align}

where $\theta_{f}$ and $\theta_{\mathcal{T}}$ represent the parameters of the model $f$ and time shifts $\mathcal{T}$ respectively. For simplicity, we assume that the time shift distribution and error distributions are independent.

@@

Before optimisation, the values of each individual time shift ($\tau$) are not known. In addition, the prediction errors $\varepsilon$ can only be calculated if each value of $\tau$ is available (because for each time shift there is a different prediction and hence prediction error). Therefore, our algorithm finds the optimum set of time shifts in an inner optimisation, while iteratively searching for the optimum parameters of $f$ in an outer optimisation loop. 
Firstly, we define a parametric form and some initial parameters to be estimated for the function $f$. For simplicity, let's take the univariate linear model where the $f(x(t))$ is parameterised by $\beta$, $\sigma_{\varepsilon}$(error standard deviation)  and $\lambda_{\tau}$ (time error parameter). First, we initialise some starting values for each of these parameters.

>>>>

\title{Modelling Stochastic Time Delay for Regression}
\author[Aaron]{Aaron Pickering}
\email{\href{mailto:aaron1rcl@gmail.com}{aaron1rcl@gmail.com}}
\author[Juan]{Juan Camilo Orduz}
\email{\href{mailto:juanitorduz@gmail.com}{juanitorduz@gmail.com}}
\urladdr{\href{https://juanitorduz.github.io/}{juanitorduz.github.io}}
%\address{}

\date{\today}
\maketitle

@@

\subsection{Model Specification}

Let $x(t)$ be a discrete time series. We want to determine the, functional relationship $f:\mathbb{R} \longrightarrow \mathbb{R}$ to some other (target) variable $y$ ob observed data $\{y_i\}_{i=1}^{n}$, that is
$y = f(x(t))$.  The series $x(t)$ must be stationary with a known value for the point at which $f(x(t)) = 0$.  That is, we know at which point the input series has no effect on the output (when it’s off). The size of the support of $f(x(t))$ should be small relative to number of points in the domain (we we will explore this condition further in the 'Limitations' section). We utilise the terminology {\em impulse} for an individual element of the support of $f$.

@@

\begin{align*}
x(t) = 
\left(
\begin{array}{cccccc}
0 & 0 & 1 & 0 & 1 & 0
\end{array}  
\right)
\end{align*}

@@

\begin{align*}
X(t) = 
\left(
\begin{array}{cccccc}
0 & 0 & 1 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 1 & 0 
\end{array}  
\right)
\end{align*}

@@

\begin{align}\label{Eqn:ModelSpecification}
y(t)= f(1^TX(t + \mathcal{T})) + \varepsilon
\end{align}

where $X(t + \mathcal{T})$ is the matrix of time shifted impulses, $X(t)$ is the observed input, $1$ is the vector $(1,1,\cdots, 1)\in 
\mathbb{R}^{1 \times k}$, where $k$ is the number of impulses and $\varepsilon$ is a noise term and $\mathcal{T}$ is the set of single impulse time delays ($\tau$'s). The model definition just asummes that the time shifts are applied to the matrix rows and the we reduce by summing up the columns. We assume that $\mathcal{T}$ is not a constant, but rather a random draw from some discrete distribution (e.g. discrete gaussian, poisson etc). For the discrete gaussian kernel $\mathcal{T}\sim T(\mu, \sigma)$ or for the poisson distribution $\mathcal{T}\sim Pois(\lambda)$. Similarly one can model the errors as  $\varepsilon \sim N(\mu, \sigma_{\varepsilon})$.

\subsection{Inference}

In order to find the best estimate of $f$, we would like to find the function $f$ which maximises the joint log-likelihood of the time-domain shifts and the prediction residuals. Specifically, we maximise:
% TODO: Clean this notation.
\begin{align}
    \mathcal{L} = 
    \underbrace{\sum_{i=1}^{n} \log(p(y_i | X_{\tau_i} = x_i; \theta_{f}, \tau_i))}_{\mathcal{L}_1} 
    + 
    \underbrace{\sum_{i=1}^{n} \log(p(\tau_i| \theta_{\mathcal{T}}))}_{\mathcal{L}_2}
\end{align}

where $\theta_{f}$ and $\theta_{\mathcal{T}}$ represent the parameters of the model $f$ and time shifts $\mathcal{T}$ respectively. In other words, the $\mathcal{L}_2$ term represents likelihood of time shifts and the $\mathcal{L}_1$ term represents the likelihood of the prediction residuals. We maximise these terms simultaneously. For simplicity, we assume that the time shift distribution and error distributions are independent.

@@

Before optimisation, the values of each individual time shift $\tau$ are not known. In addition, the prediction errors $\varepsilon$ can only be calculated if each value of $\tau$ is available (because for each time shift there is a different prediction and hence prediction error). Therefore, our algorithm finds the optimum set of time shifts in an inner optimisation, while iteratively searching for the optimum parameters of $f$ in an outer optimisation loop. 
Firstly, we define a parametric form $f(x;\theta_f)$ and some initial parameters to be estimated for the function $f$. For simplicity, let's take the univariate linear model where the $f(x(t))$ is parameterised by $\beta$, $\sigma_{\varepsilon}$(error standard deviation). In addition, we choose a poisson distribution for $\mathcal{T}$, where $\theta_{\mathcal{T}}$ is a parameterised by its mean $\lambda_{\mathcal{T}}$.

First, we initialise some starting values for each of these parameters.
